---
title: "DATA 608 Meetup 8: Visualizing Distributions"
author: "George I. Hagstrom"
format: revealjs
---

## Week Summary

- This focus is on distributions
  - Histograms, density plots, ridge plots
  - Build to working with models
- Data Story 4 due Sunday
- Reading: Chapter 7-9 of Fundamentals of Data Visualization

## Data Story 2 Review

```{r}
#| echo: false
library(tidyverse)
library(paletteer)
library(RColorBrewer)
library(ggthemes)
library(cowplot)
library(ggokabeito)
library(colorspace)
library(colorblindr)
set.seed(100)
nyc_bnb = read_csv("/home/georgehagstrom/work/Teaching/DATA608/DataStory4/nyc_airbnb_listings.csv")

neighborhood_df = nyc_bnb |> group_by(neighborhood,borough) |> summarise(num_listings = n(), mean_revenue = mean(revenue,na.rm =TRUE), mean_rating = mean(review_scores_rating,na.rm=TRUE),mean_location = mean(review_scores_location,na.rm=TRUE),mean_value = mean(review_scores_value,na.rm=TRUE)) |> filter(num_listings > 200) |> 
  ungroup() |> slice_sample(n=20) # Need ungroup to make slice_sample pick 20 neighborhoods 


library(TTR)
gdp = read_csv("../Meetup4/GDPC1.csv")
gdp = gdp |> mutate(growth = 100*(GDPC1/lag(GDPC1,4)-1)) 
unemp = read_csv("../Meetup4/UNRATE.csv")
gdp = gdp |> filter(!is.na(growth))
gdp = gdp |> left_join(unemp)


gdp |> ggplot(aes(x=observation_date,y=growth)) +
  geom_line(color="blue") +
  geom_line(aes(y=UNRATE),color="red") +
  theme_bw(base_size=16) +
  theme(panel.grid.minor.x = element_blank(),
          panel.grid.major.x = element_blank() )+
  labs(title="GDP Growth and Unemployment") +
  xlab("Date") +
  ylab("Percent")

```

## Data Story 2 Review

```{r}
gdp |> ggplot(aes(x=observation_date,y=growth)) +
  geom_line(color="blue") +
  geom_line(aes(y=UNRATE),color="red") +
  geom_line(aes(y=GDPC1),color="orange")+
  theme_bw(base_size=16) +
  theme(panel.grid.minor.x = element_blank(),
          panel.grid.major.x = element_blank() )+
  labs(title="GDP Growth and Unemployment") +
  xlab("Date") +
  ylab("Percent")


```

## Dual Axes Considerations

- Generally only plot time series on the same axis if units are comparable
- Dual axis plots very prone to misinterpretation
- See here for more [Kieran Healy Blog](https://kieranhealy.org/blog/archives/2016/01/16/two-y-axes/)


## How to Visualize Distributions

- Histogram

```{r}
nyc_bnb |> 
  ggplot(aes(price_per_accommodates)) +
  geom_histogram() +
  scale_x_continuous(limits = c(0,500)) +
  theme_minimal(base_size = 16) +
  labs(x = "Price per Accomodated Guest",
       y = "Count",
       title = "Distribution of Prices per Guest for Airbnb Accomodations in NYC ")


```

## How to Visualize Distributions

- `geom_histogram`

```{r}

#| echo: true
#| eval: false
#| 
nyc_bnb |> 
  ggplot(aes(price_per_accommodates)) +
  geom_histogram() +
  scale_x_continuous(limits = c(0,500)) +
  theme_minimal(base_size = 16) +
  labs(x = "Price per Accomodated Guest",
       y = "Count",
       title = "Distribution of Prices per Guest for Airbnb Accomodations in NYC ")


```



## How to Visualize Distributions

- Density

```{r}
nyc_bnb |> 
  ggplot(aes(price_per_accommodates)) +
  geom_density(kernel = "gaussian",fill = "skyblue",alpha=0.5) +
  scale_x_continuous(limits = c(0,500)) +
  theme_minimal(base_size = 16) +
  labs(x = "Price per Accomodated Guest",
       y = "Count",
       title = "Distribution of Prices per Guest for Airbnb Accomodations in NYC ")

```


## Histogram Binwidth

- Always experiment with binwidth

```{r}
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.3, fill = "grey50",color = "grey50") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "Bindwidth of 0.3 might be hiding features",
       x="Carat",
       y="Count")
```



## Histogram Binwidth

- Always experiment with binwidth


```{r}
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.03, fill = "grey50",color = "grey50") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "Bindwidth of 0.03 reveals customer preferences for diamonds above a threshold size",
       x="Carat",
       y="Count")
```

## Number of Bins

- Can also control number of bins

```{r}
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =10, fill = "grey50",color = "grey50") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "10 bins",
       x="Carat",
       y="Count")
```


## Number of Bins

- Can also control number of bins

```{r}
#| echo: true
#| eval: false
#| source-line-numbers: "3"
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =20, fill = "grey50",color = "grey50") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "10 bins",
       x="Carat",
       y="Count")
```


## center

- Can control centering of bins

```{r}
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =5, fill = "grey50",color = "grey50",center = 1.5) +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,4)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "Bindwidth of 0.03 reveals customer preferences for diamonds above a threshold size",
       x="Carat",
       y="Count")
```


## center

- Can control centering of bins

```{r}
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =5, fill = "grey50",color = "grey50",center = 1.5) +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,4)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "This is has a bin centered on 1.5",
       x="Carat",
       y="Count")
```



## center

- Can control centering of bins

```{r}
#| echo: true
#| eval: false

diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =5, fill = "grey50",color = "grey50",center = 1.5) +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,4)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "This is has a bin centered on 1.5",
       x="Carat",
       y="Count")
```


## boundary

- `boundary` instead of `center` sets the edge

```{r}
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =5, fill = "grey50",color = "grey50", boundary = 1) +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,4)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "Boundary of 1 forces a bin to start there",
       x="Carat",
       y="Count")
```



## boundary

- `boundary` instead of `center` sets the edge

```{r}
#| echo: true
#| eval: false
diamonds |> 
  ggplot(aes(carat)) +
  geom_histogram(bins =5, 
                 fill = "grey50",
                 color = "grey50", 
                 boundary = 1) +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,4)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "Boundary of 1 forces a bin to start there",
       x="Carat",
       y="Count")
```

## geom_freqpoly

- Histogram with lines

```{r}

diamonds |> 
  ggplot(aes(carat)) +
  geom_freqpoly(binwidth =0.03, fill = " ",color = "#CC79A7") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "",
       x="Carat",
       y="Count")
```


## geom_freqpoly

- Histogram with lines

```{r}

diamonds |> 
  ggplot(aes(carat)) +
  geom_freqpoly(binwidth =0.03, fill = " ",color = "#CC79A7") +
  geom_histogram(binwidth =0.03,fill = "#CC79A730",color="grey20") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "",
       x="Carat",
       y="Count")
```


## geom_freqpoly

- Histogram with lines

```{r}
#| echo: true
#| eval: false

diamonds |> 
  ggplot(aes(carat)) +
  geom_freqpoly(binwidth =0.03, fill = " ",color = "#CC79A7") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,3)) +
  labs(title = "Histogram of Diamond Weights",
       subtitle= "",
       x="Carat",
       y="Count")
```




## Density 

- Density estimates a smooth curve from the points
- Uses `kernel density estimation` or `kde`
$$
p(x|{x_i}) = \frac{1}{nb_W}\sum_{i=1}^n K\left(\frac{x-x_i}{b_W}\right) 
$$
- $K$ is a standard probability distribution
- $b_W$ is called the bandwidth


## Density 

- Two different bandwidths

```{r}
#| fig-width: 6
#| fig-asp: 0.67

data = tibble(x = rnorm(100))

data |> ggplot(aes(x)) +
  geom_density(bw = 0.01,color="#E69F00") +
    geom_density(bw = 0.1,color="#56B4E9") +
  theme_minimal(base_size=16) +
  labs(title="Demonstrating different bandwidth")



```


## Density 

- Two different bandwidths

```{r}
#| echo: true
#| eval: false

data = tibble(x = rnorm(100))

data |> ggplot(aes(x)) +
  geom_density(bw = 0.01,color="#E69F00") +
    geom_density(bw = 0.1,color="#56B4E9") +
  theme_minimal(base_size=16) +
  labs(title="Demonstrating different bandwidth")



```

## Kernels

- Function $K$ is a probability distribution
- Common choices controlled by `kernel` arg
  - `gaussian`, `rectangular`, `triangular`, `epanechnikov`, `biweight`, `cosine`, `optcosine`

## Kernels

- Different kernels give slightly different properties
  

```{r}

data = tibble(x = rnorm(100))

data |> ggplot(aes(x)) +
  geom_density(bw = 0.05,color="#56B4E9",fill="#56B4E950",kernel = "e") +
  theme_minimal(base_size=16) +
  labs(title="Demonstrating different Kernels",
       subtitle = "Epanechnikov is minimum RMS")

```

## Kernels

- Different kernels give slightly different properties


```{r}

data = tibble(x = rnorm(100))

data |> ggplot(aes(x)) +
  geom_density(bw = 0.05,color="#56B4E9",fill="#56B4E950",kernel = "e") +
  theme_minimal(base_size=16) +
  labs(title="Demonstrating different Kernels",
       subtitle = "Epanechnikov is minimum RMS")

```



## Kernels
  
- Different kernels give slightly different properties
  

```{r}


data = tibble(x = rnorm(100))

data |> ggplot(aes(x)) +
  geom_density(bw = 0.05,color="#56B4E9",fill="#56B4E950", kernel = "rect") +
  theme_minimal(base_size=16) +
  labs(title="Demonstrating Different Kernels",
       subtitle = "Rectangular")

```


## Kernels
  
- Different kernels give slightly different properties
  
```{r}
#| echo: true
#| eval: false
data = tibble(x = rnorm(100))

data |> ggplot(aes(x)) +
  geom_density(bw = 0.05,color="#56B4E9",fill="#56B4E950",kernel = "rect") +
  theme_minimal(base_size=16) +
  labs(title="Demonstrating Different Kernels",
       subtitle = "Rectangular")

```


## Value Cutoff

- Kernel shape can estimate unrealistic data

```{r}

nyc_bnb |> ggplot(aes(revenue)) +
  geom_density(color = palette_OkabeIto[1],fill = darken(palette_OkabeIto[1],-0.3)) +
  scale_x_continuous(limits = c(-50,1000)) +
  theme_minimal(base_size = 16) +
  labs(title = "Revenue Per Month",
       subtitle = "Why is the density non-zero for negative revenue?",
       x = "Revenue")



```


## Value Cutoff

- Use Bounds to set limits

```{r}
nyc_bnb |> ggplot(aes(revenue)) +
  geom_density(color = palette_OkabeIto[1],
               fill = darken(palette_OkabeIto[1],-0.3),
               bounds = c(0,Inf)) +
  scale_x_continuous(limits = c(-50,1000)) +
  theme_minimal(base_size = 16) +
  labs(title = "Revenue Per Month",
       subtitle = "Why is the density non-zero for negative revenue?",
       x = "Revenue")
```

## Value Cutoff

- Use Bounds to set limits

```{r}
#| echo: true
#| eval: false
nyc_bnb |> ggplot(aes(revenue)) +
  geom_density(color = palette_OkabeIto[1],
               fill = darken(palette_OkabeIto[1],-0.3),
               bounds = c(0,Inf)) +
  scale_x_continuous(limits = c(-50,1000)) +
  theme_minimal(base_size = 16) +
  labs(title = "Revenue Per Month",
       subtitle = "Why is the density non-zero for negative revenue?",
       x = "Revenue")
```




## Cumulative Distribution Function

- Alternative technique for showing a distribution
- Regular CDF:
$$
F(x) = \int_{-\infty}^x dx' p(x')
$$
- What is the probability a point drawn from $p$ is less than x?
- For data:
$$
F(x) = \frac{\# (x_i <x)}{N}
$$

## Empirical CDF


```{r}

nyc_bnb |> ggplot(aes(x = reviews_per_month,y = percent_rank(reviews_per_month))) + 
  stat_ecdf(geom = "step", color = "#0072B2") +
  geom_point(color = "#0072B2") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,25)) + 
  labs(title = "Empirical CDF of Reviews Per Month",
       y="Probability",
       x="Reviews Per Month")

```



## Empirical CDF

- `stat-ecdf`

```{r}

nyc_bnb |> ggplot(aes(x = reviews_per_month,y = percent_rank(reviews_per_month))) + 
  stat_ecdf(geom = "step", color = "#0072B2") +
  geom_point(color = "#0072B2") +
  theme_minimal(base_size = 16) +
  scale_x_continuous(limits = c(0,25)) + 
  labs(title = "Empirical CDF of Reviews Per Month",
       y="Probability",
       x="Reviews Per Month")

```




## qq-plots

- Compares ecdf to cdf of other distribution

```{r }
#| out-width: "80%"
#| fig-asp: 0.67


UScereal = MASS::UScereal

UScereal |> ggplot(aes(sample=fibre)) + 
  labs(title="QQ-Plot Fibre Content in US Cereal Brands",
       x="Theoretical Quantile",
       y="Fibre Content") +
  theme_minimal(base_size=16) +
  coord_cartesian(ylim = c(-0.1, 35)) +
  stat_qq(size = 3) + 
  stat_qq_line(color="red") 
  

```

## qq-plot

- Deviations from line indicate non-normal or problems

```{r }
#| out-width: "80%"
#| fig-asp: 0.67



UScereal = UScereal |> mutate(outlier = if_else(fibre>20,"Outlier","Normal"))
UScereal = UScereal |> mutate(inlier = if_else(fibre == 0,"Inlier","Normal"))

UScereal |> ggplot(aes(sample=fibre)) + 
  labs(title="QQ-Plot Fibre Content in US Cereal Brands",
       x="Theoretical Quantile",
       y="Fibre Content") +
  theme_minimal(base_size=16) +
  coord_cartesian(ylim = c(-0.1, 35)) +
  stat_qq(size = 3) + 
  stat_qq_line(color="red") +
  annotate("rect",xmin=1.5,xmax=2.5,ymin=25,ymax=35,,alpha = .2,linewidth=1,fill="red") +
  annotate("text",x=2,y=22,label="Potential Outliers",size = 8)
  

```

## qq-plot 

- Flat spots suggest inliers

```{r }
#| out-width: "80%"
#| fig-asp: 0.67



UScereal = UScereal |> mutate(outlier = if_else(fibre>20,"Outlier","Normal"))
UScereal = UScereal |> mutate(inlier = if_else(fibre == 0,"Inlier","Normal"))

UScereal |> ggplot(aes(sample=fibre)) + 
  labs(title="QQ-Plot Fibre Content in US Cereal Brands",
       x="Theoretical Quantile",
       y="Fibre Content") +
  theme_minimal(base_size=16) +
  coord_cartesian(ylim = c(-0.1, 35)) +
  stat_qq(size = 3) + 
  stat_qq_line(color="red") +
  annotate("rect",xmin=-2.5,xmax=-0.6,ymin=-1,ymax=2,,alpha = .2,linewidth=1,fill="orange") +
  annotate("text",x=-2,y=4,label="Potential Inliers",size = 8)
  

```


## qq-plots

- `stat_qq`
- `stat_qq_line`


```{r }
#| echo: true
#| eval: false



UScereal = UScereal |> mutate(outlier = if_else(fibre>20,"Outlier","Normal"))
UScereal = UScereal |> mutate(inlier = if_else(fibre == 0,"Inlier","Normal"))

UScereal |> ggplot(aes(sample=fibre)) + 
  labs(title="QQ-Plot Fibre Content in US Cereal Brands",
       x="Theoretical Quantile",
       y="Fibre Content") +
  theme_minimal(base_size=16) +
  coord_cartesian(ylim = c(-0.1, 35)) +
  stat_qq(size = 3) + 
  stat_qq_line(color="red") +
  annotate("rect",xmin=-2.5,xmax=-0.6,ymin=-1,ymax=2,,alpha = .2,linewidth=1,fill="orange") +
  annotate("text",x=-2,y=4,label="Potential Inliers",size = 8)
  

```



## Multiple Distributions

- Few: 
  - Small multiples
  - Same Axis with density/freq_poly if low overlap
- Slightly More?
  - Ridge Plots


## Small Multiples

```{r}

nb_vec = c("Williamsburg","West Village","Park Slope","Upper East Side")
nyc_bnb |> filter(neighborhood %in% nb_vec) |> 
  ggplot(aes(x=revenue)) +
  geom_histogram() +
  facet_wrap(~neighborhood,scales="free_y") +
  theme_minimal(base_size=16) +
  scale_x_continuous(limits=c(0,1000)) +
  labs(title="Revenue Distributions in 4 NYC Neighborhoods",
       x="Revenue")



```


## Small Multiples

```{r}
#| echo: true
#| eval: false

nb_vec = c("Williamsburg","West Village","Park Slope","Upper East Side")
nyc_bnb |> filter(neighborhood %in% nb_vec) |> 
  ggplot(aes(x=revenue)) +
  geom_histogram() +
  facet_wrap(~neighborhood,scales="free_y") +
  theme_minimal(base_size=16) +
  scale_x_continuous(limits=c(0,1000)) +
  labs(title="Revenue Distributions in 4 NYC Neighborhoods",
       x="Revenue")

```


## Multiple Densities

```{r}

nb_vec = c("Williamsburg","West Village","Park Slope","Upper East Side")
nyc_bnb |> filter(neighborhood %in% nb_vec) |> 
  ggplot(aes(x=revenue,group = neighborhood,fill = neighborhood,color=neighborhood)) +
  geom_density(alpha=0.4) +
  scale_fill_OkabeIto() +
  scale_color_OkabeIto() +
  theme_minimal(base_size=16) +
  scale_x_continuous(limits=c(0,1000)) +
  labs(title="Revenue Distributions in 4 NYC Neighborhoods",
       x="Revenue")



```

## Multiple Densities

```{r}
#| echo: true
#| eval: false
nb_vec = c("Williamsburg","West Village","Park Slope","Upper East Side")
nyc_bnb |> filter(neighborhood %in% nb_vec) |> 
  ggplot(aes(x=revenue,group = neighborhood,fill = neighborhood,color=neighborhood)) +
  geom_density(alpha=0.4) +
  scale_fill_OkabeIto() +
  scale_color_OkabeIto() +
  theme_minimal(base_size=16) +
  scale_x_continuous(limits=c(0,1000)) +
  labs(title="Revenue Distributions in 4 NYC Neighborhoods",
       x="Revenue")



```



## When to Overlap Densities

- Avoid Multiple Density Plot when there is major overlap
- OK when densities are distinct

```{r}
library(palmerpenguins)

penguins |> filter(sex == "male") |> 
  ggplot(aes(x=body_mass_g,group = species,fill = species,color=species)) +
  geom_density(alpha=0.4) +
  scale_fill_OkabeIto() +
  scale_color_OkabeIto() +
  theme_minimal(base_size=16) +
  labs(title="Male Body Mass of Penguins",
       x="Body Mass (g)")
```



## Ridge Plots

```{r}
library(ggridges)
nb_vec = c("Williamsburg","West Village","Park Slope","Upper East Side","Bedford-Stuyvesant","Astoria")
nyc_bnb |> filter(neighborhood %in% nb_vec) |> 
  ggplot(aes(x=reviews_per_month,y = neighborhood)) +
  scale_x_continuous(limits = c(0,3)) +
  geom_density_ridges(rel_min_height=0.05,alpha=0.4) +
  theme_minimal(base_size=16) +
  labs(title="Reviews Per Month in 4 NYC Neighborhoods",
       x="Reviews Per Month",y="Neighborhood")


```



## Ridge Plots

```{r}
#| echo: true
#| eval: false
library(ggridges)
nb_vec = c("Williamsburg","West Village","Park Slope","Upper East Side","Bedford-Stuyvesant","Astoria")
nyc_bnb |> filter(neighborhood %in% nb_vec) |> 
  ggplot(aes(x=reviews_per_month,y = neighborhood)) +
  scale_x_continuous(limits = c(0,3)) +
  geom_density_ridges(rel_min_height=0.05,alpha=0.4) +
  theme_minimal(base_size=16) +
  labs(title="Reviews Per Month in 4 NYC Neighborhoods",
       x="Reviews Per Month",y="Neighborhood")


```



## Thanks!



